The Blessing of FedAvg's Linear Speedup in Federated Learning

### 2020-06-01 source/sec_intro.tex

\section{Introduction}
% Motivate FL
The widespread usage of smart home devices and mobile phones constitutes
large-scale distributed networks that generate datasets of unprecedented size.
The emergence of those massive amounts of data has empowered machine learning
solutions to develop many successful applications such as personalized
recommendations~\cite{chen2018federated}, keyboard predictions~\cite{47586},
etc. Due to the increasing awareness of the privacy and rapid development of
mobile devices' computational power, it becomes more and more attractive to
push the computation to the edge. Recent years have witnessed the rise of
federated learning~\cite{mcmahan2016communication} to full fill this demand.

% Main challenge in FL.
As a distributed learning paradigm, Federated Learning (FL) collaboratively
learns a single statistical model leveraging the data from all devices while
keeping the training data isolated.  Thus, FL can integrate cross-device
information for acquiring a better model  without compromising the users'
privacy~\cite{kairouz2019advances}. Comparing to the traditional distribution optimization,  FL confronts unique
challenges~\cite{li2019convergence,smith2017federated}, including statistical heterogeneity: the data distribution in each local device can be significantly different, and system heterogeneity: only a subset of devices are accessible to the central server for model training.

% Gaps in the theoretical understanding. 
Towards resolving the aforementioned challenges, Federated
Averaging~\cite{mcmahan2016communication} (FedAvg) was proposed as an
effective heuristic and its great empirical success~\cite{47586} has attracted
lots of efforts on building up the theoretical understanding of FedAvg
and its accelerated variants. 
Roughly, the current convergence guarantee of FedAvg obtained by prior arts either established on simplified setting~\cite{stich2018local,khaled2019first,wang2018cooperative,yu2019parallel} or proved a suboptimal 
convergence rate~\cite{li2019convergence}. Furthermore, the convergence of
FedAvg's accelerated variants~\cite{yu2019linear,huo2020faster,liu2019accelerating} under the realistic setting is still waiting to be solved.

\textbf{Notations}
Throughout the paper, we adopt the following notations. Let $N$ be the total
number of local devices, and $K \leq N$ is the maximal number of devices
that are accessible to the central server.  Let $T$ be the total number of  Stochastic Gradient Descent (SGD) performed on each local device, $E$ be the local steps between two communication steps. Thus, $T/E$ is
the number of communications.

% Our contributions
\subsection{Contributions}
This work focus on providing comprehensive convergence analysis
of FedAvg and accelerated variants under the realistic setting, showing
the less number of communication round is needed to converge to the
target accuracy and preserving the linear speedup at the same time.
More specifically, for strongly convex and smooth problems, we provide the {\small{$\cO(1/NT)$}} convergence  for FedAvg and the local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. For convex and smooth objectives, we prove the {\small{$\cO(1/\sqrt{NT})$}} and the local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. The same conclusions can be equally applied for Nesterov accelerated FedAvg. To the best of our knowledge, this is the first linear speedup results for FedAvg with Nesterov accelerated SGD. Furthermore, 
we also provide the first improved convergence rate for Nesterov accelerated FedAvg comparing to FedAvg under a popular overparamterized setting. 
Last but not least, we systematically examine the convergence rate of
FedAvg and it accelerated variants on both a real-world dataset and a synthetic dataset. The empirical observations are well-aligned with theoretical indications.

@@

>>>>

\section{Introduction}
% Motivate FL
The widespread usage of smart home devices and mobile phones constitutes
large-scale distributed networks that generate amount of personal data.
Models learned from such data lead to 
% The emergence of those massive amounts of data has empowered machine learning
% solutions to develop 
many successful applications such as personalized
recommendations~\cite{chen2018federated}, keyboard predictions~\cite{47586},
and etc. Due to the rising awareness of the privacy and rapid growth of
computational power in mobile devices, it becomes increasingly attractive to
push the computation to the edge. Recent years witnessed the rise of
federated learning~\cite{mcmahan2016communication} to fulfill this demand.

% Main challenge in FL.
As a distributed learning paradigm, Federated Learning (FL) collaboratively
learns a single statistical model leveraging the data from all devices while
keeping the training data isolated.  Thus, FL integrates cross-device
information for acquiring improved models without transferring training data from individual devices, thus improves user
privacy~\cite{kairouz2019advances}. Comparing to the traditional distribution optimization,  FL confronts unique
challenges~\cite{li2019convergence,smith2017federated}, including statistical heterogeneity: the data distributions in different devices can be significantly different, and system heterogeneity: only a subset of devices may be accessible to the central server for model training at each time.

% Gaps in the theoretical understanding. 
To address the aforementioned challenges, Federated
Averaging~\cite{mcmahan2016communication} (FedAvg) was proposed as an
effective heuristic and its great empirical success~\cite{47586} attracted
lots of efforts on building up its theoretical understanding 
and accelerated variants. 
Existing convergence guarantees of FedAvg were largely established on simplified setting~\cite{stich2018local,khaled2019first,wang2018cooperative,yu2019parallel} or provides a suboptimal 
convergence rate~\cite{li2019convergence}. Furthermore, the convergence of
FedAvg's accelerated variants~\cite{yu2019linear,huo2020faster,liu2019accelerating} under the realistic setting is yet to be analyzed.

% Our contributions
% \subsection{Contributions}
\textbf{Contributions.}
This work focus on providing comprehensive convergence analysis
of FedAvg and its accelerated variants under a realistic setting.
Our results show
that the less number of communication rounds is needed to converge to a
target accuracy and preserve the linear speedup w.r.t. participating devices at the same time.
More specifically, for strongly convex and smooth problems, we show 
that FedAvg admits a convergence rate of 
{\small{$\cO(1/NT)$}} convergence, and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. For convex and smooth objectives, we show the rate of {\small{$\cO(1/\sqrt{NT})$}} and local steps can be as large as {\small{$\cO(\sqrt{T/N})$}}. 
Moreover, we show that FedAvg with Nesterov acceleration also enjoys the same rates and speedup. 
%The same conclusions are equally applied for Nesterov accelerated FedAvg. 
To the best of our knowledge, this is the first linear speedup results for FedAvg with Nesterov accelerated SGD. Furthermore, 
we provide the first improved convergence rate for Nesterov accelerated FedAvg comparing to FedAvg under the popular overparamterized setting. 
Last but not least, we systematically examine the convergence rate of
FedAvg and it accelerated variants on both a real-world dataset and a synthetic dataset. The empirical observations are well-aligned with theoretical indications.

@@

\textbf{Notations.}
Throughout the paper, we adopt the following notations. Let $N$ be the total
number of local devices, and $K \leq N$ is the maximal number of devices
that are accessible to the central server.  Let $T$ be the total number of  Stochastic Gradient Descent (SGD) iterations performed by each local device, $E$ be the local steps between two consecutive server communications. Thus, $T/E$ is
the number of communications.
